Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2023-02-02 manuscriptPNAS.Rmd
Other edits

---
title: "Different rules for binocular combination of luminance flicker in cortical and subcortical pathways"
author: "Federico G. Segala, Aurelio Bruno, Myat T. Aung, Alex R. Wade &  Daniel H. Baker"
date: "Department of Psychology, University of York, UK"

output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: yes
  pdf_document:
    toc: no
  word_document: default
  html_document: default
bibliography: references.bib
csl: pnas.csl
---

@@

```
Federico G. Segala
**Email**: fgs502@york.ac.uk

@@

**Classification:** Biological Sciences - Neuroscience/Psychological and Cognitive Sciences

@@

@@

Thirty (20 females), twelve (7 females) and ten (3 females) adult participants, whose ages ranged from 18 to 45, were recruited for Experiments 1, 2 and 3 respectively. All participants had normal or corrected to normal binocular vision, and gave written informed consent. Our procedures were approved by the Ethics Committee of the Department of Psychology at the University of York (identification number 792).

@@

The stimuli were two discs of achromatic flickering light with a diameter of 3.74 degrees, presented on a black background. The same stimuli were used for all three experiments. Four dark red lines were added around both discs to help with their fusion into one binocular disc (see insert in Figure \@ref(fig:pupildata)b for an example of the fused stimulus). The discs were viewed through a four-mirror stereoscope, which used front silvered mirrors to avoid internal reflections, and meant that participants saw a single fused disc. The use of a stereoscope allowed us to modulate the stimuli in three different ocular configurations: monocular, binocular, and dichoptic. Note that during monocular presentation of flicker, the unstimulated eye still saw the static (non-flickering) disc of mean luminance.

All stimuli had a mean luminance of 42 cd/m$^2$ on an Iiyama VisionMaster™ Pro 510 display (800 x 600 pixels, 60 Hz refresh rate), which was gamma corrected using a Minolta LS-110 photometer (Minolta Camera Co. Ltd., Japan). For experiments 1 and 2, the stimuli were presented using Psychopy (v3.0.7). For experiment 3, the stimuli were presented using Psychopy (v2022.1.1).

EEG data were collected for Experiments 1 and 2 using a 64-electrode ANT WaveGuard cap and the signals were recorded at 1 kHz using the ASA software (ANT Neuro, Netherlands). Pupillometry data were collected for Experiment 1 using a binocular Pupil Core eye-tracker (Pupil Labs GmbH, Berlin, Germany; @Kassner2014) running at 120 Hz, and the signals were recorded with the Pupil Capture software.

@@

Despite being derived from the model proposed by @Meese2006, the simplifications applied to this architecture make it very similar to other models [e.g. @Ding2006; @Legge1984; @Schrodinger1926; @Doesschate1967]. In particular we fixed the numerator exponent at 2 in our model, because otherwise this value tends to trade off with the weight of interocular suppression [see @Baker2012; @Kingdom2015]. Our key parameter of interest is the weight of interocular suppression. Large values around $w$ = 1 result in a very small or nonexistent binocular advantage at suprathreshold contrasts, consistent with previous work using grating stimuli [@Baker2017]. Low values around $w$ = 0 produce substantial, near-linear binocular facilitation [@Baker2020].

We implemented the model within a Bayesian framework using the Stan software [@Carpenter2017]. This allowed us to estimate group-level posterior parameter distributions for the weight of interocular suppression, $w$, and the other free model parameters $R_{max}$, $Z$ and $n$. The prior distributions for all parameters were Gaussian, with means and standard deviations of 1 and 0.5 for $w$ and $R_{max}$, and 5 and 2 for $Z$ and $n$, with these values chosen based on previous literature [@Meese2006; @Baker2012]. We sampled from a Student's t-distribution for the amplitudes in the pupillometry and EEG experiments, and from a Bernoulli distribution for the single trial matching data. The models were fit using the individual data across all participants, independently for each data set. Because we used coherent averaging across participants, the group average amplitudes are shifted vertically relative to the model predictions, which are based on hierarchical fits to the individual participant amplitudes (put another way, the model does not implement coherent averaging). However it is clear that the character of the model in all cases gives a good representation of the data, as can be seen in Figure \@ref(fig:modelfigure). We took posterior samples at over a million steps for each data set, using a computer cluster, and retained 10\% of samples for plotting.

@@

The pupillometry results are summarised in Figure \@ref(fig:pupildata). The group average waveform for binocular presentation is shown in Figure \@ref(fig:pupildata)a. There is a substantial pupil constriction at stimulus onset, followed by visible oscillations at the flicker frequency (2Hz, see waveform at foot). The average Fourier spectrum is displayed in Figure \@ref(fig:pupildata)b, and shows a clear spike at 2 Hz, but no evidence of a second harmonic response at 4Hz. These results demonstrate that our paradigm can evoke measurable steady-state pupil responses at 2Hz.

@@

Figure \@ref(fig:pupildata)c shows contrast response functions in response to stimuli flickering only at 2Hz. Response amplitudes increased monotonically with target contrast, confirming that our paradigm is suitable for measuring contrast-dependent differences in response (to our knowledge this is the first time this has been demonstrated). The amplitude of the binocular condition (blue squares) is consistently greater than that of the monocular condition (red circles) across all target contrasts. A $2\times5$ repeated measures $ANOVA^2_{circ}$ [@Baker2021] comparing these conditions revealed a significant main effect of target contrast (F(`r monbinPstats$dfA`,`r monbinPstats$dfR`) = `r round(monbinPstats$FratioA,digits=2)`, $p$ < 0.001), a significant effect of condition (F(`r monbinPstats$dfB`,`r monbinPstats$dfR`) = `r round(monbinPstats$FratioB,digits=2)`, $p$ < 0.001), and a significant interaction (F(`r monbinPstats$dfAB`,`r monbinPstats$dfR`) = `r round(monbinPstats$FratioAB,digits=2)`, $p$ < 0.001). The dichoptic condition begins at a much higher amplitude, owing to binocular combination of the target and high (48%) contrast mask, and then increases slightly with increasing target contrast (main effect of target contrast: F(`r dichA2c$dfM`,`r dichA2c$dfR`) = `r round(dichA2c$Fratio,digits=2)`, $p$ < `r round(dichA2c$pval,digits=3)`).

In Figure \@ref(fig:pupildata)d, we plot responses to monocular target stimuli flickering at 2Hz, when the other eye viewed stimuli flickering at 1.6Hz (the red monocular-only data are replotted from Figure \@ref(fig:pupildata)c for comparison). When the 1.6Hz component had the same contrast as the target (the binocular cross condition, shown in purple) responses were facilitated slightly at low contrasts, and suppressed at the highest target contrasts (interaction between contrast and condition: F(`r monbinPstatsX$dfAB`,`r monbinPstatsX$dfR`) = `r round(monbinPstatsX$FratioAB,digits=2)`, $p$ < 0.001). When the 1.6Hz component had a fixed contrast of 48% (the dichoptic cross condition, shown in yellow), responses were suppressed slightly across the contrast range (interaction between contrast and condition: F(`r mondichPstatsX$dfAB`,`r mondichPstatsX$dfR`) = `r round(mondichPstatsX$FratioAB,digits=2)`, $p$ < 0.001).

Figure \@ref(fig:pupildata)e shows responses at 1.6Hz, for the same conditions, as well as for a condition in which a monocular stimulus flickered at 1.6Hz (grey circles). Surprisingly there again appears to be a slight facilitation effect in the binocular cross condition, particularly at lower contrasts. The dichoptic cross condition does not show clear modulation with target contrast.

Figure \@ref(fig:EEGdata) shows equivalent results, measured contemporaneously using EEG. Figure \@ref(fig:EEGdata)a shows the group average waveform for binocular presentation, and Figure \@ref(fig:EEGdata)b shows the Fourier spectrum for binocular presentation, both averaged across four posterior electrodes (_Oz_, _POz_, _O1_ and _O2_, marked on the inset scalp plots). Unlike for the pupillometry data, there are clear responses at both the first harmonic frequency (2Hz), and also the second harmonic frequency (4Hz). We therefore calculated contrast response functions at both first and second harmonic frequencies.

@@

When stimuli in both eyes flicker at 2Hz, the binocular responses at the first (Figure \@ref(fig:EEGdata)c) and second (Figure \@ref(fig:EEGdata)f) harmonics are substantially greater than the monocular responses, particularly at high contrasts. Analysis of variance on the complex values ($ANOVA^2_{circ}$) revealed a main effect of contrast (F(`r monbinEstats$dfA`,`r monbinEstats$dfR`) = `r round(monbinEstats$FratioA,digits=2)`, $p$ < 0.001) and an interaction effect (F(`r monbinEstats$dfAB`,`r monbinEstats$dfR`) = `r round(monbinEstats$FratioAB,digits=2)`, $p$ < 0.001), but no effect of condition ($p$ = `r round(monbinEstats$pvalB,digits=2)`) at the first harmonic, with a similar pattern of results obtained at the second harmonic. For the cross-frequency conditions (Figure \@ref(fig:EEGdata)d,g), there was no appreciable effect of adding a 1.6Hz component on the response at 2Hz or 4Hz (no effect of condition, and no interaction). Similarly, there were no clear interocular interactions between frequencies in the responses at 1.6Hz (Figure \@ref(fig:EEGdata)e) and 3.2Hz (Figure \@ref(fig:EEGdata)h). This pattern of results suggests that processing of temporal luminance modulations happens in a more linear way in visual cortex (indexed by EEG), compared with subcortical pathways (indexed by pupillometry), and shows no evidence of interocular suppression.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms [e.g. @Anstis1998; @Levelt1965; @Quaia2018]. We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

The results from the matching experiment are shown in Figure \@ref(fig:matchingdata). Each data point indicates the contrast levels required in each eye that were perceptually equivalent to the binocular 24\% (red circles) and 48\% (blue circles) matching contrasts. At both matching contrasts, we see a very substantial increase in the physical contrast required for a monocular target (data points along the x- and y-axes), compared to a binocular target (points along the diagonal of x=y). For example with a 48\% match, the monocular targets required contrasts close to 100\%, whereas binocular targets required a contrast of around 50\%. The data points between these extremes also fall close to the predictions of a linear summation model (diagonal dotted lines), and are inconsistent with a winner-takes-all (or MAX) model (dashed lines). Overall, these matching results are consistent with the approximately linear summation effects observed in the EEG data of Experiment 1 (Figure \@ref(fig:EEGdata)c,f).

@@

We fitted a computational model to the data from Experiments 1 & 3 using a hierarchical Bayesian approach. The model behaviour is displayed in Figure \@ref(fig:modelfigure)e-h, with empirical data replotted in Figure \@ref(fig:modelfigure)a-d for comparison. In general, the model captures the key characteristics of the empirical data. Note that there are some minor discrepancies, which are a consequence of the hierarchical nature of the modelling. In brief, the model is fitted to the amplitudes for each participant, and group-level parameter estimates are derived based on these fits (see Table \@ref(tab:paramtable)). This procedure discards the phase information, whereas the empirical averages are coherently averaged across participants (retaining phase information). This explains the amplitude differences between model and data, particularly at low target contrast levels, but is of little consequence for the pattern of relative responses across conditions, which is our main focus here.

We were particularly interested in comparing the weight of interocular suppression across data sets. We therefore plot the posterior distributions for this parameter for all four data sets (see Figure \@ref(fig:modelfigure)i). The key finding is that the pupillometry results (green distribution) display a much greater weight of interocular suppression compared with the other data sets (grey, purple and yellow distributions). There is no overlap between the pupillometry distribution and any of the other three. All four distributions are also meaningfully below a weight of 1 -- the value that previous work using grating stimuli would predict [@Meese2006; @Baker2017], and the peak location of our prior distribution (black curve). These results offer an explanation of the empirical data: the strong interocular suppression for the pupillometry data is consistent with the weak binocular facilitation, and measurable dichoptic masking observed using that method. The weaker suppression for the other experiments is consistent with the near-linear binocular facilitation effects, and absent dichoptic masking.

@@

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e. sine wave grating stimuli) are all consistent with strong interocular suppression and weak binocular facilitation at high contrasts [@Baker2017; @Moradi2009; @Meese2006]. Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, @Quaia2018 observed a strong binocular facilitation (or 'supersummation') in the reflexive eye movement response to rapidly moving stimuli, and @Spitschan2019 report a similar result in archival data on melatonin suppression due to light exposure. In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm [@Baker2020]. Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments [@Levelt1965; @Anstis1998; @Baker2012], though not for luminance decrements [@Anstis1998]. Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of sensory signal combination.

Given the above, where does this leave our understanding of the overarching purpose of signal combination? @Baker2017 point out that strong suppression between channels that are subsequently summed is equivalent to a Kalman filter, which is the optimal method for combining two noisy inputs [see also @Ernst2002]. This account has intuitive appeal, and is consistent with other models that propose binocular combination as a means of redundancy reduction [@Li1994; @May2022]. One possibility is that optimal combination is useful for visual perception --- a critical system for interacting with the local environment --- and is therefore worth devoting the additional resource of inhibitory wiring between ocular channels. However the other examples given in the previous paragraph are primarily physiological responses (pupil size, eye movements, hormone release) that may benefit less from an increased signal-to-noise ratio, or otherwise be phylogenetically older than binocular pattern vision. Conceptualised another way, the brain can repurpose a generic architecture for different situational demands by adjusting parameter values (here the weight of interocular suppression) to achieve different outcomes. Our future work in this area intends to compare binocular combination for specific photoreceptor pathways, including different cone classes, and intrinsically photoreceptive retinal ganglion cells.

Pupil size determines the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2% of overall diameter [e.g. @Spitschan2014]. It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms [e.g. @Bradley2008]. We think it unlikely that these temporally coarse processes would have a differential effect on e.g. monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the two pathways are effectively distinct, but hope to investigate this directly in future neuroimaging work.

Classic studies investigating the neurophysiological architecture of V1 reported that cells in cytochrome-oxidase 'blobs' [@Horton1981; @Livingstone1984] are biased towards low spatial frequencies [@Tootell1988; @Edwards1995], and relatively insensitive to stimulus orientation [@Horton1981; @Livingstone1984; though see @Economides2011]. As the blob regions are embedded within ocular dominance columns [@Horton1981], they are also largely monocular [@Livingstone1984; @Tychsen2004]. More recent work has reported psychophysical evidence for unoriented chromatic [@Gheiratmand2013] and achromatic [@Meese2011] mechanisms, that also appear to be monocular. Our use of luminance flicker might preferentially stimulate these mechanisms, perhaps explaining why our EEG data show little evidence of binocular interactions. Indeed, our EEG results could potentially be explained by a model involving entirely non-interacting monocular channels, with the binocular facilitation effects we find (e.g. Figures \@ref(fig:BSratios) & \@ref(fig:TFdata)) owing to additivity of the electrophysiological response across independent monocular cells, rather than requiring binocular neurons. However our matching data (Figure \@ref(fig:matchingdata)), as well as everyday perceptual experience, indicates that luminance signals must be combined physiologically at some stage, and that this process still involves weak or absent interocular suppression.

@@

Supported by BBSRC grant BB/V007580/1 awarded to DHB and ARW, and Wellcome Trust grant 213616/Z/18/Z to AB.

>>>>

---
title: "Different rules for binocular combination of luminance flicker in cortical and subcortical pathways"
author: "Federico G. Segala, Aurelio Bruno, Myat T. Aung, Alex R. Wade &  Daniel H. Baker"
date: "Department of Psychology, University of York, UK"
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: yes
  pdf_document:
    toc: no
  word_document: default
  html_document: default
bibliography: references.bib
csl: pnas.csl
---

@@

```
Federico G. Segala, email: fgs502@york.ac.uk

@@

**Classification:** Biological Sciences - Neuroscience

@@

**This PDF includes**:

  Main Text (Abstract, Significance Statement, Introduction, Methods, Results, Discussion, Conclusions, Acknowledgments)

  Figures 1 to 6

  Table 1

@@

Thirty (20 F, 10 M), twelve (7 F, 5 M) and ten (3 F, 7 M) adult participants, whose ages ranged from 18 to 45, were recruited for Experiments 1, 2 and 3 respectively. All participants had normal or corrected-to-normal binocular vision, and gave written informed consent in accordance with the Declaration of Helsinki. Our procedures were approved by the Ethics Committee of the Department of Psychology at the University of York (identification number 792).

@@

The stimuli were two monocularly presented discs of achromatic flickering light with a diameter of 3.74 degrees, displayed against a black background. The same stimuli were used for all three experiments. Four dark red lines were added around each monocular disc to help with their fusion into a single binocular disc (see insert in Fig. \@ref(fig:pupildata)b for an example of the fused stimulus). The discs were viewed through a four-mirror stereoscope, which used front silvered mirrors to avoid internal reflections, and meant that participants perceived a single fused disc. The use of a stereoscope allowed us to modulate the stimuli in three different ocular configurations: monocular, binocular, and dichoptic. Note that during monocular presentation of flicker, the unstimulated eye still saw the static (non-flickering) disc of mean luminance.

All stimuli had a mean luminance of 42 cd/m$^2$, and were presented on an Iiyama VisionMaster™ Pro 510 display (800 x 600 pixels, 60 Hz refresh rate), which was gamma corrected using a Minolta LS-110 photometer (Minolta Camera Co. Ltd., Japan). All stimuli were presented using Psychopy [@Peirce2019] (v3.0.7 for Experiments 1 and 2; v2022.1.1 for Experiment 3).

EEG data were collected for Experiments 1 and 2 using a 64 electrode ANT WaveGuard cap and the signals were recorded at 1 kHz using the ASA software (ANT Neuro, Netherlands). Pupillometry data were collected for Experiment 1 using a binocular Pupil Core eye-tracker (Pupil Labs GmbH, Berlin, Germany @Kassner2014) running at 120 Hz, and the signals were recorded with the Pupil Capture software.

@@

Despite being derived from the model proposed by Meese et al. [@Meese2006], the simplifications applied to this architecture make it very similar to other models [@Ding2006; @Legge1984; @Schrodinger1926; @Doesschate1967]. In particular we fixed the numerator exponent at 2 in our model, because otherwise this value tends to trade off with the weight of interocular suppression [@Baker2012; @Kingdom2015]. Our key parameter of interest is the weight of interocular suppression. Large values around $w$ = 1 result in a very small or nonexistent binocular advantage at suprathreshold contrasts, consistent with previous work using grating stimuli [@Baker2017]. Low values around $w$ = 0 produce substantial, near-linear binocular facilitation [@Baker2020].

We implemented the model within a Bayesian framework using the Stan software [@Carpenter2017]. This allowed us to estimate group-level posterior parameter distributions for the weight of interocular suppression, $w$, and the other free model parameters $R_{max}$, $Z$ and $n$. The prior distributions for all parameters were Gaussian, with means and standard deviations of 1 and 0.5 for $w$ and $R_{max}$, and 5 and 2 for $Z$ and $n$, with these values chosen based on previous literature [@Meese2006; @Baker2012]. We sampled from a Student's t-distribution for the amplitudes in the pupillometry and EEG experiments, and from a Bernoulli distribution for the single trial matching data. The models were fit using the individual data across all participants, independently for each data set. Because we used coherent averaging across participants for our steady-state data, the group average amplitudes are shifted vertically relative to the model predictions, which are based on hierarchical fits to the individual participant amplitudes (put another way, the model does not implement coherent averaging). However it is clear that the character of the model in all cases gives a good representation of the data, as can be seen in Fig. \@ref(fig:modelfigure). We took posterior samples at over a million steps for each data set, using a computer cluster, and retained 10\% of samples for plotting.

@@

The pupillometry results for our first experiment are summarised in Fig. \@ref(fig:pupildata). The group average waveform for binocular presentation is shown in Fig. \@ref(fig:pupildata)a. There is a substantial pupil constriction at stimulus onset, followed by visible oscillations at the flicker frequency (2Hz, see waveform at foot). The average Fourier spectrum is displayed in Fig. \@ref(fig:pupildata)b, and shows a clear spike at 2 Hz, but no evidence of a second harmonic response at 4Hz. These results demonstrate that our paradigm can evoke measurable steady-state pupil responses at 2Hz.

@@

Fig. \@ref(fig:pupildata)c shows contrast response functions in response to stimuli flickering only at 2Hz. Response amplitudes increased monotonically with target contrast, confirming that our paradigm is suitable for measuring contrast-dependent differences in response (to our knowledge this is the first time this has been demonstrated). The amplitude of the binocular condition (blue squares) is consistently greater than that of the monocular condition (red circles) across all target contrasts. A $2\times5$ repeated measures $ANOVA^2_{circ}$ [@Baker2021] comparing these conditions revealed a significant main effect of target contrast (F(`r monbinPstats$dfA`,`r monbinPstats$dfR`) = `r round(monbinPstats$FratioA,digits=2)`, $p$ < 0.001), a significant effect of condition (F(`r monbinPstats$dfB`,`r monbinPstats$dfR`) = `r round(monbinPstats$FratioB,digits=2)`, $p$ < 0.001), and a significant interaction (F(`r monbinPstats$dfAB`,`r monbinPstats$dfR`) = `r round(monbinPstats$FratioAB,digits=2)`, $p$ < 0.001). The dichoptic condition begins at a much higher amplitude, owing to binocular combination of the target and high (48%) contrast mask, and then increases slightly with increasing target contrast (main effect of target contrast: F(`r dichA2c$dfM`,`r dichA2c$dfR`) = `r round(dichA2c$Fratio,digits=2)`, $p$ < `r round(dichA2c$pval,digits=3)`).

In Fig. \@ref(fig:pupildata)d, we plot responses to monocular target stimuli flickering at 2Hz, when the other eye viewed stimuli flickering at 1.6Hz (the red monocular-only data are replotted from Fig. \@ref(fig:pupildata)c for comparison). When the 1.6Hz component had the same contrast as the target (the binocular cross condition, shown in purple) responses were facilitated slightly at low contrasts, and suppressed at the highest target contrasts (interaction between contrast and condition: F(`r monbinPstatsX$dfAB`,`r monbinPstatsX$dfR`) = `r round(monbinPstatsX$FratioAB,digits=2)`, $p$ < 0.001). When the 1.6Hz component had a fixed contrast of 48% (the dichoptic cross condition, shown in yellow), responses were suppressed slightly across the contrast range (interaction between contrast and condition: F(`r mondichPstatsX$dfAB`,`r mondichPstatsX$dfR`) = `r round(mondichPstatsX$FratioAB,digits=2)`, $p$ < 0.001).

Fig. \@ref(fig:pupildata)e shows responses at 1.6Hz, for the same conditions, as well as for a condition in which a monocular stimulus flickered at 1.6Hz (grey circles). Surprisingly there again appears to be a slight facilitation effect in the binocular cross condition, particularly at lower contrasts. The dichoptic cross condition does not show clear modulation with target contrast.

Fig. \@ref(fig:EEGdata) shows equivalent results, measured contemporaneously using EEG. Fig. \@ref(fig:EEGdata)a shows the group average waveform for binocular presentation, and Fig. \@ref(fig:EEGdata)b shows the Fourier spectrum for binocular presentation, both averaged across four posterior electrodes (_Oz_, _POz_, _O1_ and _O2_, marked on the inset scalp plots). Unlike for the pupillometry data, there are clear responses at both the first harmonic frequency (2Hz), and also the second harmonic frequency (4Hz). We therefore derived contrast response functions at both first and second harmonic frequencies.

@@

When stimuli in both eyes flicker at 2Hz, the binocular responses at the first (Fig. \@ref(fig:EEGdata)c) and second (Fig. \@ref(fig:EEGdata)f) harmonics are substantially greater than the monocular responses, particularly at high contrasts. Analysis of variance on the complex values ($ANOVA^2_{circ}$) revealed a main effect of contrast (F(`r monbinEstats$dfA`,`r monbinEstats$dfR`) = `r round(monbinEstats$FratioA,digits=2)`, $p$ < 0.001) and an interaction effect (F(`r monbinEstats$dfAB`,`r monbinEstats$dfR`) = `r round(monbinEstats$FratioAB,digits=2)`, $p$ < 0.001), but no effect of condition ($p$ = `r round(monbinEstats$pvalB,digits=2)`) at the first harmonic, with a similar pattern of results obtained at the second harmonic (note that the absence of a main effect of condition is expected, given that there should be negligible differences between conditions at low contrasts, and any differences are therefore captured by the interaction term). For the cross-frequency conditions (Fig. \@ref(fig:EEGdata)d,g), there was no appreciable effect of adding a 1.6Hz component on the response at 2Hz or 4Hz (no effect of condition, and no interaction). Similarly, there were no clear interocular interactions between frequencies in the responses at 1.6Hz (Fig. \@ref(fig:EEGdata)e) and 3.2Hz (Fig. \@ref(fig:EEGdata)h). This pattern of results suggests that processing of temporal luminance modulations happens in a more linear way in visual cortex (indexed by EEG), compared with subcortical pathways (indexed by pupillometry), and shows no evidence of interocular suppression.

@@

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms [@Anstis1998; @Levelt1965; @Quaia2018]. We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

The results from the matching experiment are shown in Fig. \@ref(fig:matchingdata). Each data point indicates the contrast levels required in each eye that were perceptually equivalent to the binocular 24\% (red circles) and 48\% (blue circles) matching contrasts. At both matching contrasts, we see a very substantial increase in the physical contrast required for a monocular target (data points along the x- and y-axes), compared to a binocular target (points along the diagonal of x=y). For example with a 48\% match, the monocular targets required contrasts close to 100\%, whereas binocular targets required a contrast of around 50\%. The data points between these extremes also fall close to the predictions of a linear summation model (diagonal dotted lines), and are inconsistent with a winner-takes-all (or MAX) model (dashed lines). Overall, these matching results are consistent with the approximately linear summation effects observed in the EEG data of Experiment 1 (Fig. \@ref(fig:EEGdata)c,f).

@@

We fitted a computational model to the data from Experiments 1 & 3 using a hierarchical Bayesian approach. The model behaviour is displayed in Fig. \@ref(fig:modelfigure)e-h, with empirical data replotted in Fig. \@ref(fig:modelfigure)a-d for comparison. In general, the model captures the key characteristics of the empirical data. Note that there are some minor discrepancies, which are a consequence of the hierarchical nature of the modelling. In brief, the model is fitted to the amplitudes for each participant, and group-level parameter estimates are derived based on these fits (see Table \@ref(tab:paramtable)). This procedure discards the phase information, whereas the empirical averages are coherently averaged across participants (retaining phase information). This explains the amplitude differences between model and data, particularly at low target contrast levels, but is of little consequence for the pattern of relative responses across conditions, which is our main focus here.

We were particularly interested in comparing the weight of interocular suppression across data sets. We therefore plot the posterior distributions for this parameter for all four data sets (see Fig. \@ref(fig:modelfigure)i). The key finding is that the pupillometry results (green distribution) display a much greater weight of interocular suppression compared with the other data sets (grey, purple and yellow distributions). There is no overlap between the pupillometry distribution and any of the other three. All four distributions are also meaningfully below a weight of 1 -- the value that previous work using grating stimuli would predict [@Meese2006; @Baker2017], and the peak location of our prior distribution (black curve). These results offer an explanation of the empirical data: the strong interocular suppression for the pupillometry data is consistent with the weak binocular facilitation, and significant dichoptic masking observed using that method. The weaker suppression for the other experiments is consistent with the near-linear binocular facilitation effects, and absent dichoptic masking.

@@

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e. sine wave grating stimuli) are universally consistent with strong interocular suppression and weak binocular facilitation at high contrasts [@Baker2017; @Moradi2009; @Meese2006]. Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, Quaia et al.@Quaia2018 observed a strong binocular facilitation (or 'supersummation') in the reflexive eye movement response to rapidly moving stimuli, and Spitschan and Cajochen@Spitschan2019 report a similar result in archival data on melatonin suppression due to light exposure. In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm [@Baker2020]. Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments [@Levelt1965; @Anstis1998; @Baker2012], though not for luminance decrements [@Anstis1998]. Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of binocular signal combination.

Given the above, where does this leave our understanding of the overarching purpose of signal combination? We previously@Baker2017 pointed out that a model with strong suppression between channels that are subsequently summed is equivalent to a Kalman filter, which is the optimal method for combining two noisy inputs (see also [@Ernst2002]). This account has intuitive appeal, and is consistent with other models that propose binocular combination as a means of redundancy reduction [@Li1994; @May2022]. One possibility is that optimal combination is useful for visual perception --- a critical system for interacting with the local environment --- and is therefore worth devoting the additional resource of inhibitory wiring between ocular channels. However the other examples given in the previous paragraph are primarily physiological responses (pupil size, eye movements, hormone release) that may benefit less from an increased signal-to-noise ratio, or are otherwise phylogenetically older than binocular pattern vision. Conceptualised another way, the brain can repurpose a generic architecture for different situational demands by adjusting parameter values (here the weight of interocular suppression) to achieve different outcomes. Our future work in this area intends to compare binocular combination for specific photoreceptor pathways, including different cone classes, and intrinsically photoreceptive retinal ganglion cells.

Pupil size determines the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2% of overall diameter [@Spitschan2014]. It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms [@Bradley2008]. We think it unlikely that these temporally coarse processes would have a differential effect on e.g. monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the two pathways are effectively distinct, but intend to investigate this directly in future neuroimaging work.

Classic studies investigating the neurophysiological architecture of V1 reported that cells in cytochrome-oxidase 'blobs' [@Horton1981; @Livingstone1984] are biased towards low spatial frequencies [@Tootell1988; @Edwards1995], and relatively insensitive to stimulus orientation [@Horton1981; @Livingstone1984] (though see @Economides2011). As the blob regions are embedded within ocular dominance columns [@Horton1981], they are also largely monocular [@Livingstone1984; @Tychsen2004]. More recent work has reported psychophysical evidence for unoriented chromatic [@Gheiratmand2013] and achromatic [@Meese2011] mechanisms, that also appear to be monocular. Our use of luminance flicker might preferentially stimulate these mechanisms, perhaps explaining why our EEG data show little evidence of binocular interactions. Indeed, our EEG results could potentially be explained by a model involving entirely non-interacting monocular channels, with the binocular facilitation effects we find (e.g. Fig. \@ref(fig:BSratios) & \@ref(fig:TFdata)) owing to additivity of the electrophysiological response across independent monocular cells, rather than requiring binocular neurons. However our matching data (Fig. \@ref(fig:matchingdata)), as well as everyday perceptual experience, indicates that luminance signals must be combined physiologically at some stage, and that this process still involves weak or absent interocular suppression.

@@

Supported by Biotechnology and Biological Sciences Research Council grant BB/V007580/1 awarded to DHB and ARW, and Wellcome Trust grant 213616/Z/18/Z to AB.
